---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 2 - Due date 01/23/25"
author: "Jingze Dai"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(openxlsx)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
energy_data <- read.xlsx(
  xlsxFile="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
  sheet = "Monthly Data", startRow = 13, colNames = FALSE)

read_col_names  <- read.xlsx(
  xlsxFile="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
  sheet = "Monthly Data", rows = 11, colNames = FALSE)

colnames(energy_data) <- read_col_names
energy_data$Data <- convertToDate(energy_data$Month)
head(energy_data)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
# selecting the columns of interest
power_data <- energy_data %>% 
  select("Total Biomass Energy Production",
         "Total Renewable Energy Production", 
         "Hydroelectric Power Consumption")

head(power_data)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts_power_data <- ts(power_data, start=c(1973,1), frequency=12)
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
mean_ts <- sapply(ts_power_data, mean)
sd_ts <- sapply(ts_power_data, sd)
print("Mean for the three series")
mean_ts
print("Standard deviation for the three series")
sd_ts
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r fig.height=10, fig.width=6}

par(mfrow = c(3,1))
plot(ts_power_data[,1],
     type="l",col="blue",
     ylab="Total Biomass Energy Production (Trillion Btu)",
     main="Time Series for Biomass Energy Production") 
abline(h=mean(ts_power_data[,1]),col="red")

plot(ts_power_data[,2],
     type="l",col="green",
     ylab="Total Renewable Energy Production (Trillion Btu)",
     main="Time Series for Renewable Energy Production") 
abline(h=mean(ts_power_data[,2]),col="red")

plot(ts_power_data[,3],
     type="l",col="purple",
     ylab="Hydroelectric Power Consumption (Trillion Btu)",
     main="Time Series for Hydroelectric Power Consumption") 
abline(h=mean(ts_power_data[,3]),col="red")
```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_power_data)
```

Only total biomass energy and renewable energy productions are positively correlated, as their correlation coefficient is 0.968, close to 1. Hydroelectric power consumption is not correlated to any of the other two variables since the correlation coefficients are close to 0, at -0.114 and -0.029 respectively.

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r fig.height=8, fig.width=5}

par(mfrow = c(3,1))
biomass_acf<-acf(ts_power_data[,1],lag.max=40, 
                main = "ACF for Total Biomass Energy Production",
                type="correlation", plot=TRUE,
                xlab="Lag (years)")
renewable_acf<-acf(ts_power_data[,2],lag.max=40, 
                  main = "ACF for Total Renewable Energy Production",
                  type="correlation", plot=TRUE,
                  xlab="Lag (years)")
hydro_acf<-acf(ts_power_data[,3],lag.max=40, 
              main = "ACF: Total Hydroelectric Consumption",
              type="correlation", plot=TRUE,
              xlab="Lag (years)")
```
For both the biomass and renewable energy consumption, the ACF values decrease gradually as lags increase, indicating a strong trend. However, it does not have the sinusoidal pattern, suggesting that no significant seasonality is present.

For the hydroelectric consumption, the ACF plot shows a sinusoidal pattern with a periodicity of 12 lags, indicating a seasonal behavior. However, the positive sinusoidal wave has stronger autocorrelation than the negative side.

*Also, ideally we only want to add lags that are significant to feed into the model, thus we want initial significance at the start of the lags.*

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r fig.height=8, fig.width=5}
par(mfrow = c(3,1))
biomass_pacf<-pacf(ts_power_data[,1],lag.max=40, 
                main = "PACF for Total Biomass Energy Production",
                plot=TRUE,xlab="Lag (years)")
renewable_pacf<-pacf(ts_power_data[,2],lag.max=40, 
                  main = "PACF for Total Renewable Energy Production",
                  plot=TRUE,xlab="Lag (years)")
hydro_pacf<-pacf(ts_power_data[,3],lag.max=40, 
              main = "PACF: Total Hydroelectric Consumption",
              plot=TRUE,xlab="Lag (years)")
```

All PACF plots shows seasonality. For total hydroelectric consumption, the seasonality is expected as the shape of the PACF plot is similar to its ACF plot. However, for the other two time series, after removing the effects of other lags, there is significant ACF values at the lag value of 12 months (or 1 year), which is different from the ACF plot.

*Also there is dependence on lag 1 for the first two time series because after removing the effect of the first lag the PACF of other lags decreased.*